Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-07 TOMS_Paper/assembly.tex
Rewrite zero-columns avoidance

\subsection{Avoiding Iteration on Zero-blocks by Symbolic Execution}
\label{sec:zeros}
The second task of the ER consists of skipping arithmetic operations over blocks of zero-valued entries in basis function and derivatives of basis functions arrays. Zero columns in such arrays arise, for example, when taking derivatives on a reference element and when employing mixed elements. In~\cite{quadrature1}, a code transformation to eliminate floating point operations on zero columns was introduced; the technique was implemented in FEniCS. We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}. We propose a strategy in which we avoid the use of indirection arrays in the generated code (e.g. \texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}), which otherwise would break the optimizations applicable at the Code Specializer level, especially SIMD vectorization.

@@

Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the transformed Burgers excerpt in Figure~\ref{fig:factorized-code}. The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh. The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points. We notice the presence of 4 zero-columns. Any multiplications or additions along these columns could (should) be skipped in order to minimize irrelevant floating point operations. One solution, as anticipated, is not to generate the zero columns, i.e. to generate a dense 6$\times$2 array, to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g. $ind = \lbrace 3, 5\rbrace$) to access the right entries in the element tensor $A$. As explained, this prevents, among the various optimizations, effective code vectorizability, because memory loads and store would eventually reference non-contiguous locations. 
%on the other hand, reshrinki tutto su un array solo...

Our approach is based on domain knowledge and symbolic execution. We discern the origin of zero-columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements. In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to preserve the correctness of the element matrix evaluation while iterating along the space of test and trial functions. The two key observations are that 1) the number of zero-columns caused by mixed elements is much larger then that due to derivatives, and 2) such columns are contiguous in memory. Based on this, we decide to skip any computation involving zero-columns induced by mixed elements, whereas we still iterate over any other zero-columns. Consequently, the only modifications to the code that are eventually needed are adjusting loop bounds and introducing offsets when accessing the element matrix, as shown in Figure~\ref{fig:withzeros-skipped-code}. In general, multiple iteration spaces over test and trial functions (i.e. \texttt{jk} loops), characterized by their own loop bounds, may be needed, increasing loop overhead and decreasing data locality for the element matrix.

The implementation is based on symbolic execution. The ER expects information from the higher layer about where the zero-blocks are located, in each tabulated basis function. This could come in the form of code annotation if the input to COFFEE were provided as a C string (as shown in Figure~\ref{fig:withzeros-skipped-code}) or by suitably decorating basis functions nodes in the abstract syntax tree representing the local assembly routine. Then, each statement in the local assembly loop nest is symbolically executed. For example, consider the statement \texttt{T2[r] = ((d*D[i][k])+(e*E[i][k]));} in Figure~\ref{fig:withzeros-skipped-code}. Array \texttt{D} has non-zeros in positions $NZ_D = [3,5]$; we also assume \texttt{E} has non-zeros in positions $NZ_E = [0,2]$. Multiplications by scalar do not affect the propagation of zero columns. On the other hand, when summing the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we record that the target identifier \texttt{T2} will have non-zero columns in positions $NZ_D \cup NZ_E = [0-5]$. Exploiting the $NZ$ information associated with each identifier, the assembly expression is split into several sets of sub-expressions, each set characterized by accesses to the same range of non-zero columns and, therefore, iterating over a suitably-sized iteration space. In our example, there is just one set, which contains two sub-expressions, enclosed in a doubly nested loop of size 3$\times$3.

...TODO...: dire come si riaggiustano gli indici quando si vuole usare AVX ma la parte dense non e' multiplo del vector length - se non e' paddata la tail, allora si decrementa lo starting index di un pochino, tanto sono zeri...

\section{Code Specialization}
\label{sec:code-spec}
The Code Specializer is the second software component of COFFEE. It provides a range of code transformations tailored for optimizing instruction-level parallelism and register locality. As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying. In this paper, we enrich this set of transformations as detailed in the following sections.

>>>>

\subsection{Avoiding Iteration on Zero-blocks by Symbolic Execution}
\label{sec:zeros}
We aim at skipping arithmetic operations over blocks of zero-valued entries in basis functions (and their derivatives) arrays. Zero-valued columns arise, for example, when taking derivatives on a reference element and when employing mixed elements. In~\cite{quadrature1}, a technique to avoid operations on zero-valued columns based on the use of indirection arrays was described (e.g. \texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}) and implemented in the context of FEniCS. We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}. Essentially, our strategy avoids indirection arrays in the generated code, which otherwise would break the optimizations applicable at the Code Specializer level, including SIMD vectorization.

@@

Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}. The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh. The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points. There are four zero-valued columns. Any multiplications or additions along these columns could (should) be skipped to avoid irrelevant floating point operations. The solution adopted in~\cite{quadrature1} is not to generate the zero-valued columns (i.e. to generate a dense 6$\times$2 array), to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g. $ind = \lbrace 3, 5\rbrace$) to update the right entries in the element tensor $A$. This prevents, among the various optimizations, effective SIMD vectorization, because memory loads and store would eventually reference non-contiguous locations. 
%on the other hand, reshrinki tutto su un array solo...

Our approach is based on using domain knowledge and symbolic execution. We discern the origin of zero-valued columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements. In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to correctly evaluate the local element matrix while iterating along the space of test and trial functions. The two key observations are that 1) the number of zero-valued columns caused by using vector function spaces is, often, much larger then that due to derivatives, and 2) such columns are contiguous in memory. Based on this, we aim at avoiding iteration only along the block of zero-valued columns induced by mixed (vector) elements.

%Our example is then transformed as in Figure~\ref{fig:withzeros-skipped-code}: loop bounds are adjusted and suitable offsets are introduce to access the element matrix and basis function arrays. In general, the element matrix evaluation may also have to be split over multiple iteration spaces (test and trial functions loops), each iteration space characterized by its own loop bounds; this has the side effect of both increasing loop overhead and decreasing data locality.

We achieve our goal by symbolic execution. The Expression Rewriter expects some indication in the input exposing the location of the zero-valued columns due to mixed (vector) function spaces, in each tabulated basis function. This indication comes either in the form of code annotation, if the input to COFFEE were provided as pure C (as shown in Figure~\ref{fig:withzeros-code}), or by suitably decorating basis functions nodes, if the input were an abstract syntax tree. Then, the rewrite rules are applied and each statement is executed symbolically. For example, consider the assignment \texttt{T2[r] = d*D[i][k]+e*E[i][k]} in Figure~\ref{fig:withzeros-code}. Array \texttt{D} has non-zero-valued columns in positions $NZ_D = [3,5]$; we also assume array \texttt{E} has non-zero-valued columns in positions $NZ_E = [0,2]$. Multiplications by scalar values do not affect the propagation of non-zero-valued columns. On the other hand, when symbolically executing the sum of the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we track that the target identifier \texttt{T2} will have non-zero-valued columns in positions $NZ_D \cup NZ_E = [0,5]$. Eventually, exploiting the $NZ$ information computed and associated with each identifier, we split the original assembly expression into multiple sets of sub-expressions, each set characterized by the same range of non-zero-valued columns. In our example, assuming that $NZ_{T1} = [3,5]$ and $NZ_A = [3,5]$, there are two of such sets, this leads to the generation of two distinct iteration spaces, one for each set, as in Figure~\ref{fig:withzeros-skipped-code}.

\section{Code Specialization}
\label{sec:code-spec}
Code specialization provides a range of code transformations tailored to optimizing for instruction-level parallelism and register locality. As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying. In this paper, we enrich this set of transformations as explained in the following sections.

\subsection{Padding and Data Alignment Revisited}
Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided. We recall effective SIMD vectorization can be achieved only if the innermost loop size is a multiple of the vector register length. In the case of an AVX instruction set, for example, we want the size of innermost loops to be multiples of 4 (vector registers can fit up to four double-precision floats). Moreover, efficient loads and stores can be issued only if all of the involved arrays' base addresses are multiple of the vector register length.

Consider again the code in Figure~\ref{fig:withzeros-skipped-code}. The arrays in the loop nest $[j1,k1]$ can be padded and the right bound of loop \texttt{k1} can be safely increased to 8: eventually, values computed in the region $M[0:3][6:8]$ will be discarded. Then, by explicitly aligning arrays and using suitable pragmas (e.g. \texttt{$\#$pragma simd} for the Intel compiler), effective SIMD auto-vectorization can be obtained.

There are some complications in the cose of loops $[j0,k0]$. Here, increasing the loop bound to 4, despite being still safe, has no effect: for instance, the starting addresses of load instructions are \texttt{T1[3]} and \texttt{A[i][3]}, which are not aligned. One solution is to start iterating from the closest index that would ensure data alignment: in this specific case, $k0=0$, which would mean losing the effect of expression rewriting. Another possibility is to attain to non-aligned accesses. COFFEE can generate code for both situations, so we leave the autotuner, described in Section~\ref{sec:autotune}, in charge of determining the optimal transformation.

Note that in some cases it is not even safe to increase loop bounds, since extra iterations would access non-zero entries in the local element matrix $M$. In such a situation, there may be no possibility of restoring data alignment at all.
